---
format: gfm
---

# Understanding seasonal patterns of acute malnutrition and common childhood diseases in Afghanistan

## Insights from 14 of acute malnutrition and five years of childhood illnesses admission data
👋 Hello, and welcome to this repository!

This repository presents a time series analysis conducted to glean insights into the seasonal variations of acute malnutrition and common childhood illnesses in Afghanistan. The analysis draws on a rich dataset, spanning 14 years of acute malnutrition admissions and five years of admissions for common child illnesses. 

The ultimate objective is to generate evidence-based insights to support the IPC Acute Malnutrition analysis parameters and to guide the optimal timing of nutrition data collection efforts in the country.

::: {.callout-note}

This analysis was made possible thanks to the strong collaboration of Afghanistan’s Nutrition Cluster Coordination Team and UNICEF. Their support and insights were instrumental throughout the process. 

:::

## A glance at the results 
### The time series plot
```{r}
#| label: source-project
#| eval: true
#| echo: false
Sys.setenv(path_secret_key = "~/.ssh/id_rsa")
suppressMessages(source(file = "script.R"))
```
```{r}
#| label: time-plot
#| fig-width: 17
#| fig-height: 8
#| echo: false
tsplot +
    labs(
        title = "Evolution of acute malnutrition admissions over time: from January 2013 to December 2024",
        subtitle = "The level of admissions rose as of September 2021 like never seen before and remained high thereafter"
    )
```

### The components 

```{r}
#| label: components
#| fig-height: 8
#| fig-width: 16
#| echo: false
cmpnts_ntnl_plot
```

## Repository Structure
The repository is structured in the following way:

  + `data/raw`: data.frames containing admissions over time. Data is reported on a monthly basis, with a reporting rate over 98%. The reporting rate is defined as the number of catchment areas that submitted their reported in a given month, divided by the overall number of catchment areas that are expected to report. 
  + `R/`: some handy user-defined functions for the project.  
  + `scripts/`: A set of `R` scripts used for the analysis.

The project workflow can be implemented simply by running the `script.R` file found in the root directory.

## Reproducibility information
The repository was created in `R` version 4.5.1 This project uses the `{renv}` framework to record `R` package dependencies and versions. Packages and versions used are recorded in `renv.lock` and code used to manage dependencies is in `renv/` and other files in the root project directory. On starting an `R` session in the working directory, run `renv::restore()` to install R package dependencies.

## Data encryption  
This project uses `{cyphr}` to encrypt the raw data that lives in `data-raw/` directory. In order to be able to access and decrypt the encrypted data, the user will need to have created their own personal SSH key and make a request to be added to the project. An easy-to-grasp guide on how to make a request will be found [here](https://github.com/OxfordIHTM/cyphr-encryption-demonstration#)

## License
This repository is licensed under a GNU General Public License 3 (GPL-3).  

## Feedback
If you wish to give feedback, file an issue or seek support, kindly do so [here](https://github.com/nutimes/afg-amn-admit/issues).

## Author
Tomás Zaba
