---
title: "Understanding seasonal patterns of acute malnutrition and common childhood diseases in Afghanistan"
subtitle: "Evidence from 14 and five years of admission data on acute malnutrition and diseases, respectively"
---

```{r}
#| label: source-project
Sys.setenv(path_secret_key = "~/.ssh/id_rsa")
suppressMessages(source(file = "script.R"))
```

## Background

::: {.incremental}
* Seasonality/seasonal variation refers to recurring patterns at specific times of the year. 
&nbsp;
* It has impact on the underlying and immediate factors that drives acute malnutrition (AMN). 
&nbsp;
* It is known that AMN has a seasonal variation, however limited evidence is available to substantiate that.  
  + Stakeholders often rely on assumptions of an AMN seasonal peak during the food security lean season.  
    + Nutrition surveys are routinely timed to coincide with the lean season each year. 
::: 

## Objective

* Understand the seasonal patterns of acute malnutrition and the most common childhood diseases, in order to provide evidence-based inputs for IPC AMN analysis parameters and to inform the timing of nutrition data collection in Afghanistan

## Methods {.smaller}
:::: {.columns}
::: {.column style="text-align: left"}
#### Source of data

* **Fourteen years** of admission data on acute malnutrition into the country's treatment programme
    + Data was provided: 
        - Broken down by severe acute malnutrition (SAM) and moderate acute malnutrition cases
            - These were aggregated to form the Global Acute Malnutriton (GAM)
    + Broken down by provinces (all of them)

* Five years of admission data on child diseases
:::
::: {.column style="text-align: left"}

#### Analysis

* **Exploratory (visual) data analysis**: to inspect possible underneath patterns
    + If: 
        - patterns were similar across provinces: 
            - aggregate data as National
    + Otherwise: 
        - Analyse the provinces with distinct patterns as one

* **Time series decomposition**: a **process** of **extracting** the underlying **trend** and **seasonal variation** from the time series

* Current seasonal variation was determined on the basis of the patterns exihibited over the last three years  

:::
::::

# Results
* **Acute Malnutrition**

## {.smaller}
### A glance at the evolution of admissions over time (1/2)

```{r}
#| label: time-plot
#| fig-width: 17
#| fig-height: 8
tsplot
```

## {.smaller}
### A glance at the evolution of admissions over time (2/2)
```{r}
#| label: time-plot-year
#| fig-width: 12
#| fig-height: 4.3
tsplot_byear
```

#### Key messages

+ The level of admissions rose as of September 2021, like never seen before. It reached its highest point in December 2021 with **over 150K cases**. Thereafter, it remained high over time, like never seen before as well. 

::: {.callout-important}
## Question?
What may have hapenned in the 3rd quarter of 2021 and thereafter?

+ Insights from the Nutrition Cluster Coordination Team, following the presentation of preliminary findings, suggest that this is due to the scale-up of programme operations. With changes in government, previously inaccessible areas have become accessible for interventions.
:::

## 
### The components

```{r}
#| label: components
#| fig-height: 8
#| fig-width: 16
cmpnts_ntnl_plot
```

## 
### Seasonal patters - National 
```{r}
#| label: seasonal-national
#| fig-width: 17
#| fig-height: 8
ss_ntnl
```

## {.smaller}
### Evolution of the seasonal pattern over time
```{r}
#| label: evolution-seasonality-national
#| fig-width: 12
#| fig-height: 4.8
#| fig-subcap: 
#|   - "Before 2015"
#|   - "Between 2016 and 2019"
#|   - "Between 2020 and 2022"
#|   - "As of 2022"
#| layout-nrow: 2
#| layout-ncol: 2
ss_ntnl_b2015
ss_ntnl_2016_2019
ss_ntnl_2020_2022
ss_ntnl_asof2022
```

## {.smaller}
### Seasonal patters in Farah and Nimrus Provinces (1/2)
```{r}
#| label: seasonal-nimrus-farah
#| fig-width: 17
#| fig-height: 8
nimurs_farah
```

## {.smaller}
### Evolution of the seasonal pattern over time in Farah and Nimrus Provinces 
```{r}
#| label: evolution-seasonality-nimfarh
#| fig-width: 12
#| fig-height: 5.8
#| fig-subcap: 
#|   - "Before 2015"
#|   - "Between 2016 and 2017"
#|   - "Between 2018 and 2019"
#|   - "Between 2020 and 2021"
#|   - "As of 2021"
#| layout-nrow: 2
#| layout-ncol: 3
nimurs_farah_b2015
nimurs_farah2016_2017
nimurs_farah2018_2019
nimurs_farah_2020_2021
nimurs_farah_asof2021
```

# Translating graphs into seasonal calendars 🗓️

## {.smaller}
### AMN seasonal calendar 
::: {.gantt-seasonality-amn}
```{mermaid}
%%| label: amn-seasonal-calendar
%%| fig-cap-location: bottom
%%| fig-height: 5
%%| fig-width: 14
gantt
    title  AMN seasonal calendar based on the pattens as of 2022
    dateFormat M
    axisFormat %b
    todayMarker off

    section National
    Nadir :1, 6
    Peak : 6, 7
    Nadir: 7, 8
    High peak :8, 10
    Nadir :10, 12

    section Farah and Nimruz
    Nadir :1, 6
    Peak : 6, 7
    Nadir: 7, 9
    High peak :9, 11
    Nadir :11, 12

```
:::

* **Key message**
    + 2 High AMN seasons
        - In Farah and Nimruz Provinces, the high-peak exhibits a slight shift compared to the national
    + 1 Long low AMN season

# Results
* **Common childhood diseases**

## {.smaller}
### A glance at the evolution of admissions over time

```{r}
#| label: time-plot-disease
#| fig-width: 17
#| fig-height: 8
morbi_ntln |> 
    autoplot()
```

## {.smaller}
### The components

```{r}
#| label: components-morbi
#| fig-width: 17
#| fig-height: 8
cmpnts_morbi_ntnl_plot
```

## {.smaller}
### Seasonal patters - National 
```{r}
#| label: seasonal-national-morbi
#| fig-width: 17
#| fig-height: 8
seasonal_morbi_ntln
```

# Translating graphs into seasonal calendars 🗓️

## {.smaller}
### Child morbidity seasonal calendar 

<br> 

::: {.gantt-seasonality-morbi}
```{mermaid}
%%| label: morbi-seasonal-calendar
%%| fig-cap-location: bottom
%%| fig-height: 5
%%| fig-width: 14
gantt
    title  Child morbidity seasonal calendar
    dateFormat M
    axisFormat %b
    todayMarker off

    section Acute Respiratory Infections
    Stable (high) :1, 3
    Decrease : 3, 7
    Increase: 7, 12

    section Pneumonia
    Stable (high) :1, 3
    Decrease : 3, 7
    Increase: 7, 12

    section AWD
    Decrease : 1, 3
    Increase : 3, 8
    Decrease : 8, 12

    section Malaria
    Decrease : 1, 4
    Increase : 4, 9
    Decrease : 9, 12

    section Measles
    Decrease : 1, 2
    Increase : 2, 4
    Decrease : 4, 12

```
:::


## {.smaller}
### AMN, child morbidity, AFI and access constraint seasons
::: {.gantt-seasonality-amn-factors}
```{mermaid}
%%| label: amn-factors-seasonal-calendar
%%| fig-cap-location: bottom
%%| fig-height: 7
%%| fig-width: 14

gantt
    dateFormat M
    axisFormat %b
    todayMarker off

    section National AMN
    Nadir :1, 6
    Peak : 6, 7
    Nadir: 7, 8
    High peak :8, 10
    Nadir :10, 12

    section Farah and Nimruz's AMN
    Nadir :1, 6
    Peak : 6, 7
    Nadir: 7, 9
    High peak :9, 11
    Nadir :11, 12

    section Acute Respiratory Infections
    Stable (high) :1, 3
    Decrease : 3, 7
    Increase: 7, 12

    section Pneumonia
    Stable (high) :1, 3
    Decrease : 3, 7
    Increase: 7, 12

    section AWD
    Decrease : 1, 3
    Increase : 3, 8
    Decrease : 8, 12

    section Malaria
    Decrease : 1, 4
    Increase : 4, 9
    Decrease : 9, 12

    section Measles
    Decrease : 1, 2
    Increase : 2, 4
    Decrease : 4, 12

    section Acute Food Insecurity
    Lean : 1, 4
    Harvest : 4, 7
    Post-harvest : 7, 11
    Lean: 11, 12

    section Season
    Winter : 1, 3
    Spring: 3, 6
    Summer: 6, 9
    Autum: 9, 11
    Winter: 11, 12

    section Access constraints
    Low access in mountainous and northern regions: 1, 3
    Gradual improvement in access : 3, 5
    Generally better access: 5, 7
    Good physical access, but high insecurity: 7, 9
    Mixed access conditions: 9, 11
    Declining access in many regions: 11, 12

```
:::